A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2020-03-03 paper/main.tex
minor typos

\begin{abstract}
The Iterated Prisoner's Dilemma has been used for decades as a model of
behavioural interactions. From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of strategies in the game for years. The results of the literature,
however, have been relying on a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments,
casting doubt on the generalizability of results from statistical perspective.
This manuscript evaluates \numberofstrategies strategies in more than 40000
tournaments, presents the top performing strategies, and analyzes their
salient features.
The outcomes support that a strategy with a theory of mind should adjust it’s
cooperations based on the mean/median cooperator of its environment to be
successful; the adjustments are distinct for different tournament types.
\end{abstract}

@@

Conceptualising strategies and understanding the best way of playing the game
has been of interest to the scientific community since the formulation of the
game in 1950~\cite{Flood1958}. Following the computer tournaments of Axelrod in the
1980's~\cite{Axelrod1980a, Axelrod1980b}, a strategy's performance in a round
robin computer tournament became a common evaluation technique for newly designed
strategies. Today more than 200 strategies exist in the literature and several
tournaments, excluding Axelrod's, have been undertaken~\cite{Bendor1991,
Harper2017, Kendall2007, Stephens2002, Stewart2012}.

In the 80's, Axelrod performed two computer tournaments~\cite{Axelrod1980a,
Axelrod1980b}. The contestants were strategies submitted in the form of computer
code. They competed against all other entries, a copy of themselves and a random
strategy. The winner was decided on the average score a strategy achieved. The
winner of both tournaments was the simple strategy Tit For Tat which cooperated
on the first turn and then simply copied the previous action of it's opponent.
Due to the strategy's strong performance in both tournaments, and moreover in a
series of evolutionary experiments~\cite{Axelrod1981}, Tit For Tat was thought
to be the most robust basic strategy in the IPD.

@@

In 2004, a $20^{\text{th}}$ Anniversary Iterated Prisoner Dilemma Tournament
took place with 233 entries. This time the winning strategy was not designed on
a reciprocity based approach but on a mechanism of
teams~\cite{J.P.Delahaye1993Lp, J.P.Delahaye1995LIeP, A.Rogers2007Ctpw}. A team
from Southampton University took advantage of the fact that a participant was
allowed to submit multiple strategies. They submitted a total of 60 strategies
that could recognised each other and colluded to increase one members score.
This resulted with three of the strategies to be ranked in the top spots. The
performance of the Southampton University team received mixed attention, though
they had won the tournament as stated in~\cite{us_blog} "technically this
strategy violates the spirit of the Prisoner's Dilemma, which assumes that the
two prisoners cannot communicate with one another".

Another set of IPD strategies that have received a lot of attention
are the zero-determinant strategies (ZDs)~\cite{Press2012}. By
forcing a linear relationship between the payoffs ZDs can ensure that they will
never receive less than their opponents. The
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire''.
ZDs are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their simplicity and extortionate
behaviour have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including ZDs, was ran and none of them
ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based on lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

Though only a select pieces of work have been discussed, the IPD literature is
rich, and new strategies and competitions are being published every year~\cite{Glynatsi2019}. The
question, however, still remains the same: what is the best way to play the
game? Compared to other works, whereas a few selected strategies are evaluated
on a small number of tournaments, this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. These tournaments do not
consist by just standard round robin tournaments, but also by tournaments with noise
and tournaments with a probabilistic ending. The later part of the paper, evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The data set used
in this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.

@@

In comparison, the top ranked strategies in noisy environments when \(p_n\in [0,
1]\) include strategies which decide their actions based on the cooperations to
defections ratio, such as ShortMem~\cite{Andre2013},
Grumpy~\cite{axelrodproject} and e~\cite{axelrodproject}, and the Retaliate
strategies which are designed to defect if the opponent has tricked them more
often than x\% of the times that they have done the same. The overall
performance of these strategies is bimodal,
Figure~\ref{fig:noise_overall_results}. This indicates that although the top
ranked strategies mainly performed well, there are several tournaments that they
ranked in the bottom half. The bimodality of the \(r\) distributions can be
explained by Figure~\ref{fig:effect_of_noise} which gives the \(r\)
distributions for the top 6 strategies over the noise probability \(p_n\).
It is evident that the highly ranked strategies
were highly ranked due to the their performance in tournaments with \(p_n>0.5\),
and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly.

@@

Up till now, the performances of the \numberofstrategies strategies have been
evaluated for individual tournament types. The distributions of \(r\) for the
tournament types indicate that for standard, probabilistic ending and noisy
tournaments with \(p_n <0.1\) successful strategies do exist. For these settings
the top 15 strategies frequently ranked in the top spots with only a few
exceptions. The behaviour of the winning strategies in probabilistic ending
tournaments varied when \(p_e\) was restricted to 0.1. For larger values of
\(p_e\) the dominant strategies were defecting ones and for smaller values the
dominant strategies were evolved strategies and strategies based on teams.
In noisy tournaments Similarly the behaviour
of the dominant strategies varied when the probability of noisy was restricted
to. In noise tournaments with \(p_n < 0.1\) the dominant strategies were strategies
specifically trained for noise tournaments, and the performed well in most of the
tournaments they participated. However, in noisy tournaments with \(p_n \in [0, 1]\)
the performances of the top ranked strategies were bimodal. Indicating that with
high values of noisy a successful set of strategies does not exist.

@@

The top ranks include strategies that have been previously mentioned. The
set of Retaliate strategies occupy the top spots followed by BackStabber
and DoubleCrosser. The distributions of the Retaliate strategies have no
statistical difference. Thus, in an IPD tournament where the type is not
specified, playing as any of the Retaliate strategies will have the result.
DoubleCrosser performed well in standard tournaments and the
strategy is just an extension of BackStabber. It should be noted that these
strategies can be characterised as "cheaters". The source code of the strategies
allows them to known the number of turns in a match (if they are specified).
PSO Gambler and Evolved HMM 5 are
trained strategies introduced in~\cite{Harper2017} and Nice Meta Winner and NMWE
Memory One are strategies based on teams. Grudger is a strategy from Axelrod's
original tournament and Forgetful Fool Me Once is based on the same approach as
Grudger. Overall the top 15 strategies are fundamentally different. Some are cheaters,
some are complex, others are simple deterministic strategies and strategies based
on teams. The results of \numberofalltournaments tournaments used in this work imply the following:
they is not a single type of strategy performs well in any IPD interaction.

@@

APL makes use of classifiers to classify strategies according to
various dimensions. These determine whether
a strategy is stochastic or deterministic, whether it makes use of the number of
turns or the game's payoffs. The memory usage feature is calculated as the
memory size of strategy (which is specified in the strategies implementation
in the APL) divide by the number of turns. For example, Evolved
FSM 16 Noise 05 has a memory size of 16 and participated in a tournament where
$n$ was 134. In the given tournament Evolved FSM 16 Noise 05 has a memory usage
of 0.119. If the memory size of a strategy is infinite then memory usage is 1.
For tournaments with a probabilistic ending the number of turns was
not collected, so the memory usage feature is not used for probabilistic ending
tournaments. The SSE is a feature introduced in~\cite{Knight2019} which shows
how close a strategy is to behaving as a ZDs, and subsequently, in an
extortionate way. The method identifies the ZDs closest to a given strategy and
calculates the algebraic distance between them, defined as SSE. A SSE value of 1
indicates no extortionate behaviour at all whereas a value of 0 indicates that a
strategy is behaving a ZDs. The rest of the features considered are the $CC$ to
$C$, $CD$ to $C$, $DC$ to $C$, and $DD$ to $C$ rates as well as cooperating
ratio of a strategy. The minimum, maximum, medium and median cooperating ratios
of each tournament are also included, and finally the number of turns, the
number of strategies, the number of repetitions and the probabilities of noise
and the game ending are also included.

@@

Compared to standard tournaments, in both noisy and in probabilistic ending
tournaments the higher the rates of cooperation the lower a strategy's success
and median score. A strategy would want to cooperate less than both
the mean and median cooperator in such settings. In probabilistic ending
tournaments the correlations coefficients have a larger values, indicating a
stronger effect. Thus a strategy will be punished more by its cooperative
behaviour in probabilistic ending environments, this was seen in Section~\ref{section:evaluation_of_performance}
as well. The distributions of the $C_r$ of the winners in
both tournaments is given by Figure~\ref{fig:c_r_distributions}. It confirms
that the winners in noisy tournaments cooperated less than 35\% of the times
and in probabilistic ending tournaments less than 10\%.
In noisy probabilistic ending tournaments and in over all the tournaments' results,
the only features that had a moderate affect are $C_r/C_{\text{mean}},
C_r/C_{\text{max}}$ and $C_r$. In such environments cooperative behaviour
appears to be punished by not as much as in noisy and probabilistic ending
tournaments.

@@

This manuscript has explored the performance of \numberofstrategies strategies of the Iterated
Prisoner's Dilemma in a large number of computer tournaments. The results of
the analysis demonstrated that, although for specific tournament types such as
standard, probabilistic ending tournaments and noisy tournaments with \(p_n <0.1\), dominant types of strategies exist
there is not a single dominant strategy in an IPD competition if the environments
vary. Moreover, a strategy with a theory of mind should aim to adapt its behaviour
based on the mean and median cooperators and should in general not be too
cooperative.

The \numberofstrategies strategies used in this manuscript have been mainly for
the literature, and they have been accessible due to an open source software
called the Axelrod-Python library. The software was used to generate a total of
\numberofalltournaments computer tournaments results with different number of
strategies and different participants each time. The data collection was
described in Section~\ref{section:data_collection}. In
Section~\ref{section:top_performances}, the tournaments results were used to
present the top performances. The data set contained results from four different
settings, and these were also studied individually. In standard tournaments
complex strategies trained using reinforcement learning ranked in the top spots.
Some of these strategies ranked again in the top spots in probabilistic
ending tournaments when a \(p_e\) of less 0.1 was considered and in noisy tournaments
when \(p_n\) was less than 0.1. In probabilistic
ending tournaments \(p_e\) was designed to vary between 0 and 1. It was demonstrated
that for values larger than 0.1, as stated in the Folk Theorem, defecting strategies
were winning the tournaments because there was a high likelihood of the game
ending in the next turn. In tournaments with noise the median ranks of the top
15 strategies had the highest values and the \(r\) distributions were bimodal.
The top rank strategies were performing both well and bad, and this indicates
that in noisy tournaments there are not strategies that can guarantee winning.
However, if the probability of noisy was constrained at 0.1 then strategies
designed for noisy tournaments indeed performed well.
Overall, the top ranked strategies differed from one tournament type to
another and the mechanism behind the winning strategies were all different, with some
exceptions which include trained (Evolved) and Meta strategies.

>>>>

\begin{abstract}
The Iterated Prisoner's Dilemma has been used for decades as a model of
behavioural interactions. From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of strategies in the game for decades. The results of the literature,
however, have been relying on a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments,
casting doubt on the generalizability of results from statistical perspective.
This manuscript evaluates \numberofstrategies strategies in more than 40000
tournaments, presents the top performing strategies, and analyzes their
salient features.
The outcomes support that a strategy with a theory of mind should adjust its
cooperations based on the mean/median cooperator of its environment to be
successful; the adjustments are distinct for different tournament types.
\end{abstract}

@@

Conceptualising strategies and understanding the best way of playing the game
has been of interest to the scientific community since the formulation of the
game in 1950~\cite{Flood1958}. Following the computer tournaments of Axelrod in the
1980's~\cite{Axelrod1980a, Axelrod1980b}, a strategy's performance in a round
robin computer tournament became a common evaluation technique for newly designed
strategies. Many tournaments have followed Axelrod's~\cite{Bendor1991,
Harper2017, Kendall2007, Stephens2002, Stewart2012} and today more than 200
strategies exist in the literature.

In the 80's, Axelrod performed two computer tournaments~\cite{Axelrod1980a,
Axelrod1980b}. The contestants were strategies submitted in the form of computer
code. They competed against all other entries, a copy of themselves and a random
strategy. The winner was decided on the average score a strategy achieved. The
winner of both tournaments was the simple strategy Tit For Tat which cooperated
on the first turn and then simply copied the previous action of its opponent.
Due to the strategy's strong performance in both tournaments, and moreover in a
series of evolutionary experiments~\cite{Axelrod1981}, Tit For Tat was often
claimed to be the most robust basic strategy in the IPD.

@@

In 2004, a $20^{\text{th}}$ Anniversary Iterated Prisoner Dilemma Tournament
took place with 233 entries. This time the winning strategy was not designed on
a reciprocity based approach but on a mechanism of
teams~\cite{J.P.Delahaye1993Lp, J.P.Delahaye1995LIeP, A.Rogers2007Ctpw}. A team
from Southampton University took advantage of the fact that a participant was
allowed to submit multiple strategies. They submitted a total of 60 strategies
that could recognise each other and collude to increase one members score.
This resulted in three of the strategies ranking in the top spots. The
performance of the Southampton University team received mixed attention, though
they had won the tournament as stated in~\cite{us_blog} "technically this
strategy violates the spirit of the Prisoner's Dilemma, which assumes that the
two prisoners cannot communicate with one another".

Another set of IPD strategies that have received a lot of attention
are the zero-determinant strategies (ZDs)~\cite{Press2012}. By
forcing a linear relationship between the payoffs ZDs can ensure that they will
never receive less than their opponents. The
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire''.
ZDs are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their superiority in tournament settings
have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including ZDs, was ran and none of them
ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based on lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

Though only select pieces of work have been discussed, the IPD literature is
rich, and new strategies and competitions are being published every year~\cite{Glynatsi2019}. The
question, however, still remains the same: what is the best way to play the
game? Compared to other works, whereas a few selected strategies are evaluated
on a small number of tournaments, this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. These tournaments come in a
number of variations including tournaments with noise, probabilistic match
length and both noise and probabilistic match length.
The later part of the paper, evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The data set used
in this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.

@@

In comparison, the top ranked strategies in noisy environments when \(p_n\in [0,
1]\) include strategies which decide their actions based on the cooperations to
defections ratio, such as ShortMem~\cite{Andre2013},
Grumpy~\cite{axelrodproject} and e~\cite{axelrodproject}, and the Retaliate
strategies which are designed to defect if the opponent has tricked them more
often than x\% of the times that they have done the same. The overall
performance of these strategies is bimodal,
Figure~\ref{fig:noise_overall_results}, indicating that although the top
ranked strategies mainly performed well, there are several tournaments that they
ranked in the bottom half. The bimodality of the \(r\) distributions can be
explained by Figure~\ref{fig:effect_of_noise} which gives the \(r\)
distributions for the top 6 strategies over the noise probability \(p_n\).
It is evident that the highly ranked strategies
were highly ranked due to the their performance in tournaments with \(p_n>0.5\),
and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly.

@@

Up till now, the performances of the \numberofstrategies strategies have been
evaluated for individual tournament types. The distributions of \(r\) for the
tournament types indicate that for standard, probabilistic ending and noisy
tournaments with \(p_n <0.1\) successful strategies do exist. For these settings
the top 15 strategies frequently ranked in the top spots with only a few
exceptions. The behaviour of the winning strategies in probabilistic ending
tournaments varied when \(p_e\) was restricted to 0.1. For larger values of
\(p_e\) the dominant strategies were defecting ones and for smaller values the
dominant strategies were evolved strategies and strategies based on teams.
Similarly the behaviour
of the dominant strategies varied when the probability of noisy was restricted.
In noisy tournaments with \(p_n < 0.1\) the dominant strategies were strategies
specifically trained for noise tournaments, and the performed well in most of the
tournaments they participated. However, in noisy tournaments with \(p_n \in [0, 1]\)
the performances of the top ranked strategies were bimodal. Indicating that with
high values of noisy a successful set of strategies does not exist.

@@

The top ranks include strategies that have been previously mentioned. The
set of Retaliate strategies occupy the top spots followed by BackStabber
and DoubleCrosser. The distributions of the Retaliate strategies have no
statistical difference. Thus, in an IPD tournament where the type is not
specified, playing as any of the Retaliate strategies will have same the result.
DoubleCrosser performed well in standard tournaments and the
strategy is just an extension of BackStabber. It should be noted that these
strategies can be characterised as "cheaters". The source code of the strategies
allows them to known the number of turns in a match (if they are specified).
PSO Gambler and Evolved HMM 5 are
trained strategies introduced in~\cite{Harper2017} and Nice Meta Winner and NMWE
Memory One are strategies based on teams. Grudger is a strategy from Axelrod's
original tournament and Forgetful Fool Me Once is based on the same approach as
Grudger. Overall the top 15 strategies are fundamentally different. Some are cheaters,
some are complex, others are simple deterministic strategies and strategies based
on teams. The results of \numberofalltournaments tournaments used in this work imply the following:
they is not a single type of strategy performs well in any IPD interaction.

@@

APL classifies strategies along several dimensions. These determine whether
a strategy is stochastic or deterministic, whether it makes use of the number of
turns or the game's payoffs. The memory usage feature is calculated as the
memory size of strategy (which is specified in the strategies implementation
in the APL) divide by the number of turns. For example, Evolved
FSM 16 Noise 05 has a memory size of 16 and participated in a tournament where
$n$ was 134. In the given tournament Evolved FSM 16 Noise 05 has a memory usage
of 0.119. If the memory size of a strategy is infinite then memory usage is 1.
For tournaments with a probabilistic ending the number of turns was
not collected, so the memory usage feature is not used for probabilistic ending
tournaments. The SSE is a feature introduced in~\cite{Knight2019} which shows
how close a strategy is to behaving as a ZDs, and subsequently, in an
extortionate way. The method identifies the ZDs closest to a given strategy and
calculates the algebraic distance between them, defined as SSE. A SSE value of 1
indicates no extortionate behaviour at all whereas a value of 0 indicates that a
strategy is behaving a ZDs. The rest of the features considered are the $CC$ to
$C$, $CD$ to $C$, $DC$ to $C$, and $DD$ to $C$ rates as well as cooperating
ratio of a strategy. The minimum, maximum, medium and median cooperating ratios
of each tournament are also included, and finally the number of turns, the
number of strategies, the number of repetitions and the probabilities of noise
and the game ending are also included.

@@

Compared to standard tournaments, in both noisy and in probabilistic ending
tournaments the higher the rates of cooperation the lower a strategy's success
and median score. A strategy would want to cooperate less than both
the mean and median cooperator in such settings. In probabilistic ending
tournaments the correlation coefficients have a larger values, indicating a
stronger effect. Thus a strategy will be punished more by its cooperative
behaviour in probabilistic ending environments, this was seen in Section~\ref{section:evaluation_of_performance}
as well. The distributions of the $C_r$ of the winners in
both tournaments is given by Figure~\ref{fig:c_r_distributions}. It confirms
that the winners in noisy tournaments cooperated less than 35\% of the times
and in probabilistic ending tournaments less than 10\%.
In noisy probabilistic ending tournaments and in over all the tournaments' results,
the only features that had a moderate affect are $C_r/C_{\text{mean}},
C_r/C_{\text{max}}$ and $C_r$. In such environments cooperative behaviour
appears to be punished by not as much as in noisy and probabilistic ending
tournaments.

@@

This manuscript has explored the performance of \numberofstrategies strategies of the Iterated
Prisoner's Dilemma in a large number of computer tournaments. The results of
the analysis demonstrated that, although for specific tournament types such as
standard, probabilistic ending tournaments and noisy tournaments with \(p_n <0.1\), dominant types of strategies exist
there is not a single dominant strategy, from the collection on strategies considered in this work, in an IPD competition if the environments
vary. Moreover, a strategy with a theory of mind should aim to adapt its behaviour
based on the mean and median cooperators and should in general not be too
cooperative.

The \numberofstrategies strategies used in this manuscript have been mainly for
the literature, and they have been accessible due to an open source software
called the Axelrod-Python library. The software was used to generate a total of
\numberofalltournaments computer tournaments results with different number of
strategies and different participants each time. The data collection was
described in Section~\ref{section:data_collection}. In
Section~\ref{section:top_performances}, the tournaments results were used to
present the top performances. The data set contained results from four different
settings, and these were also studied individually. In standard tournaments
complex strategies trained using reinforcement learning ranked in the top spots.
Some of these strategies ranked again in the top spots in probabilistic
ending tournaments when a \(p_e\) of less 0.1 was considered and in noisy tournaments
when \(p_n\) was less than 0.1. In probabilistic
ending tournaments \(p_e\) was designed to vary between 0 and 1. It was demonstrated
that for values larger than 0.1, as stated in the Folk Theorem, defecting strategies
were winning the tournaments because there was a high likelihood of the game
ending in the next turn. In tournaments with noise the median ranks of the top
15 strategies had the highest values and the \(r\) distributions were bimodal.
The top rank strategies were performing both well and bad, and this indicates
that in noisy tournaments there are not strategies that can guarantee winning.
However, if the probability of noise was constrained at 0.1 then strategies
designed for noisy tournaments indeed performed well.
Overall, the top ranked strategies differed from one tournament type to
another and the mechanism behind the winning strategies were all different, with some
exceptions which include trained (Evolved) and Meta strategies.
